---
title: "STAT/MATH 495: Problem Set 09"
author: "Leonard Yoon"
date: "2017-11-07"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    collapsed: false
    smooth_scroll: false
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.width=8, fig.height=4.5, message=FALSE, warning = FALSE
  )
set.seed(76)

library(tidyverse)
library(proxy)
```



# Collaboration

Please indicate who you collaborated with on this assignment: 



# Question 1: Run k-means

```{r}
observations_1 <- read_csv("data/observations_1.csv")
observations_2 <- read_csv("data/observations_2.csv")

# Set observations to be one of two datasets
observations <- observations_1

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

**Questions**:

1. Run KMC 10 times on `observations_1` and comment on the consistency of the
results.
1. Speculate on the root cause of any consistency or inconsistency in the
results.
1. Run KMC 10 times on `observations_2` and comment on the consistency of the
results.
1. Speculate on the root cause of any consistency or inconsistency in the
results.

**Answers**:

1. The results are rather inconsistent in that there are 4 possible outcomes: cluster 1 is approximately the top half of points, and cluster 2 is the bottom half of points; cluster 1 is approximately the left half of points, and cluster 2 is the right half of points; and vice versa to both scenarios. However, all of the points that are in the left/right cluster are always in the left/right cluster in every iteration (no flip-flop of points in the middle), and same for the top/bottom clusters.
1. The inconsistency comes from the fact that the initialization step in `kmeans` entails random assignment of cluster to all points, and the centers of these random points might be farther away vertically or horizaontally. If the initial centers are closer vertically but farther away horizontally, it becomes more likely that after optimization, cluster 1 will be the left or right half of points and cluster 2 will be the other half. If the initial centers are closer horizontally but father away vertically, it becomes more likely that after optimization, cluster 1 will be the top or bottom half of points and cluster 2 will be the other half.
1. The results are consistent in that all of the points in a given cluster are always in the cluster together, though the designation of cluster 1 and cluster 2 seems to be random. That is, the points in the top right quadrant are always cluster 1 or cluster 2 and the points in the bottom left quadrant are the other cluster (I'm defining the "quadrant" as left or right of x1 = 1.0 and above or below x2 = 1.0).
1. The inconsistency comes from the fact that the initialization step entails random assignments of clusters to all points, and the centers of these random points might be closer to one quadrant or another. For example, the center of cluster 1 might happen to initialize closer to the top right quadrant for one iteration of `kmeans` and after optimization, the center of cluster 1 becomes the center of the points in that upper right quadrant. But one can easily imagine a similar scenario where the center of cluster 1 happens to initialize closer to the bottom left quadrant in another iteration of `kmeans`. My results are consistent once I run the model because there is one approximate line of symmetry in `observations_2` (with an equation $y = 2 - x$) that serves a "border" between the two clusters. That is, if the initialized point is above that line of symmetry, it will eventually (after iterations) become the center of the cluster in the top right corner. If the initialized point is below that line of symmetry, it will eventually become the center of the cluster in the bottom left corner.



# Bonus question: Code your own

Read ISLR page 388 Algorithm 10.1 and implement k-means clustering from scratch.
Don't worry about doing it for general $k$; keep it simple and do it for $k=2$
specifically. Apply it to `observations_2` from above.

```{r}
# Hint:
A <- data_frame(
  x1 = c(0, 0.5, 0.75, 1),
  x2 = c(0, 0.5, 0.75, 1)
)
B <- data_frame(
  x1 = c(1, 0),
  x2 = c(1, 0)
)
distance_matrix <- proxy::dist(x=A, y=B)
distance_matrix
apply(distance_matrix, 1, which.min) # chooses the column within each row containing the min
apply(distance_matrix, 2, which.min) # chooses the row within each column containing the min
```

In the distance matrix above, the number in [1,1] corresponds to the distance of (0,0) from (1,1) (which is $\sqrt{2}$). The number in [1,2] corresponds to the distance of (0,0) from (0,0) (which is 0).

```{r, eval=FALSE, include=FALSE}
observations_2 <- observations_2 %>% 
  mutate(ID = rownames(observations_2)) %>% # create ID
  select(ID, x1, x2) # reorder columns, ID first
A <- observations_2 %>% # randomly choose points in first cluster
  sample_n(50)
B <- observations_2 %>% 
  anti_join(A, by="ID")
center <- data_frame( # 2 random centers with coordinates (x1, y1) and (x2, y2)
  x1 = c(runif(1, min = range(observations_2$x1)[1], max = range(observations_2$x1)[2]), # x1
         runif(1, min = range(observations_2$x2)[1], max = range(observations_2$x2)[2])), # x2
  x2 = c(runif(1, min = range(observations_2$x1)[1], max = range(observations_2$x1)[2]), # y1
         runif(1, min = range(observations_2$x2)[1], max = range(observations_2$x2)[2])) # y2
)
distance_matrix_A <- proxy::dist(x=A[-1], y=center) # ignores ID for purposes of calculating distance
distance_matrix_A
distance_matrix_B <- proxy::dist(x=B[-1], y=center)
distance_matrix_B
A <- A %>% mutate(cluster = apply(distance_matrix_A, 1, which.min))
apply(distance_matrix_B, 1, which.min)

# I'm stuck trying to reassign points from 1 or 2 into a new A and new B
```
